Concurrent Programming in Harmony

### 2020-12-27 book/paper.tex
Latest
got rid of badsize
Added ~

\chapter{List of Values}
\label{ap:values}

@@

Tuples, lists, strings, and bags are all special cases of dictionaries.
Both tuples and lists map indexes (starting at 0) to Harmony values.
Their format is either \texttt{($e$, $e$, ..., $e$,)} or
\texttt{[$e$, $e$, ..., $e$,]}.
If the tuple or list has two or more elements, then the final comma
is optional.
A string is represented as a tuple of its characters.
Characters are one-character atoms, which can be expressed
in hexadecimal unicode using the syntax \texttt{.0xXX}.
A bag or multiset is a dictionary that maps a value to how many
times it occurs in the bag.

@@

@@

{\small
\begin{tabular}{|l|l|}
\hline
\texttt{$e$ == $e$}, \texttt{$e$ != $e$} & two Harmony values of the same type \\
\hline
\texttt{$e$ < $e$, $e$ <= $e$, $e$ > $e$, $e$ >= $e$} & any two Harmony values \\
\hline
\texttt{$e$ and $e$ and ...} & two or more booleans \\
\hline
\texttt{$e$ or $e$ or ...} & two or more booleans \\
\hline
\texttt{$e$ => $e$} & two booleans (implication) \\
\hline
\texttt{not $e$} & a boolean \\
\hline
\texttt{-$e$, \string~$e$} & an integer \\
\hline
\texttt{$e$ + $e$ + ...} & two or more integers, tuples, or lists \\
\hline
\texttt{$e$ - $e$} & two integers or sets \\
\hline
\texttt{$e$ \& $e$ \& ...} & two or more integers or sets \\
\hline
\texttt{$e$ | $e$ | ...} & two or more integers or sets \\
\hline
\texttt{$e$ \string^ $e$ \string^ ...} & two or more integers or sets \\
\hline
\texttt{$e$ * $e$} & two integers or an integer and a list \\
\hline
\texttt{$e$ / $e$, $e$ // $e$} & two integers (integer division) \\
\hline
\texttt{$e$ \% $e$, $e$ mod $e$} & two integers (division remainder) \\
\hline
\texttt{$e$ ** $e$, $e << e$, $e >> e$} & two integers \\
\hline
\texttt{$\{ e..e \}$} & two integers \\
\hline
\texttt{$e$ [not] in $e$} & first is any Harmony value, second is a set or dict \\
\hline
\texttt{$e$ if $e$ else $e$} & middle $e$ must be a boolean \\
\hline
\texttt{!$e$} & an address \\
\hline
\texttt{choose $e$} & a set \\
\hline
\texttt{min $e$, max $e$, any $e$, all $e$, len $e$} & a set or a dict \\
\hline
\texttt{keys $e$} & a dictionary (includes tuple and list) \\
\hline
\texttt{bagsize $e$} & a bag \\
\hline
\texttt{atLabel $e$} & atom (corresponding to a label) \\
\hline
\texttt{nametag $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{threads $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{setintlevel} \textit{e} & \textit{e} is a boolean \\
\hline
\texttt{stop} \textit{lv} & \textit{lv} is an lvalue \\
\hline
\texttt{?}\textit{lv} & \textit{lv} is an lvalue \\
\hline
\texttt{lambda $a$: $e$ end} & $a$ is a bounded variable, $e$ an expression\\
\hline
\end{tabular}
}
\vspace{1em}

@@

\section{The \texttt{bag} module}

@@

\section{The \texttt{set} module}

>>>>

\chapter{Value Types}
\label{ap:values}

@@

Tuples, lists, strings, and bags are all special cases of dictionaries.
Both tuples and lists map indexes (starting at 0) to Harmony values.
Their format is either \texttt{($e$, $e$, ..., $e$,)} or
\texttt{[$e$, $e$, ..., $e$,]}.
If the tuple or list has two or more elements, then the final comma
is optional.
A string is represented as a tuple of its characters.
Characters are one-character atoms, which can be expressed
in hexadecimal unicode using the syntax \texttt{.0xXX}.
A bag or multiset is a dictionary that maps a value to how many
times it occurs in the bag.
The syntax is \texttt{bag\{ $e$, $e$, ... \}}.

@@

Generic operators on Harmony values include:

\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{$e$ == $e$} & equivalence \\
\hline
\texttt{$e$ != $e$} & inequivalence \\
\hline
\texttt{$e$ < $e$, $e$ <= $e$, $e$ > $e$, $e$ >= $e$} & comparison\\
\hline
\end{tabular}
\end{center}

\section{Booleans}

The Boolean type has only two possible values: \texttt{False} and
\texttt{True}.  Unlike in Python, in Harmony booleans are distinct
from integers, and in particular $\mathtt{False} < 0$.  In statements
and expressions where booleans are expected, it is not possible to
substibute values of other types.

Operations on booleans include:
\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{$e$ and $e$ and ...} & conjuction \\
\hline
\texttt{$e$ or $e$ or ...} & disjunction \\
\hline
\texttt{$e$ => $e$, $e$ not => $e$} & implication \\
\hline
\texttt{not $e$} & negation \\
\hline
\texttt{$v$ if $e$ else $v'$} & $v$ or $v'$ depending on $e$ \\
\hline
\texttt{any $s$, all $s$} & disjunction / conjunction for set $s$ \\
\hline
\end{tabular}
\end{center}

\section{Integers}

The integer type supports any whole number, as well as $-\mathtt{inf}$
and $\mathtt{inf}$.  In the Python-based model checker, integers are
infinite precision.  In the C-based model checker, integers are implemented
by two's complement 61-bit words, and $-\mathtt{inf}$ is represented by
the minimal integer and $\mathtt{inf}$ is represented by the maximal integer.

Operations on integers include:

\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{-$e$} & negation \\
\hline
\texttt{abs($e$)} & absolute value \\
\hline
\texttt{$e$ + $e$ + ...} & sum \\
\hline
\texttt{$e$ - $e$} & difference \\
\hline
\texttt{$e$ * $e$ * $e$ ...} & product \\
\hline
\texttt{$e$ / $e$, $e$ // $e$} & integer division \\
\hline
\texttt{$e$ \% $e$, $e$ mod $e$} & integer division remainder \\
\hline
\texttt{$e$ ** $e$} & power \\
\hline
\texttt{\string~$e$} & binary inversion \\
\hline
\texttt{$e$ \& $e$ \& ...} & binary and \\
\hline
\texttt{$e$ | $e$ | ...} & binary or \\
\hline
\texttt{$e$ \string^ $e$ \string^ ...} & binary exclusive or \\
\hline
\texttt{$e << e$} & binary shift left \\
\hline
\texttt{$e >> e$} & binary shift right \\
\hline
\texttt{$\{ e..e' \}$} & set of integers from $e$ to $e'$ inclusive \\
\hline
\end{tabular}
\end{center}

\section{Atoms}

Atoms are essentially names and consist of one or more unicode characters.
If they do not contain special characters and do not start with a digit,
then an atom can be represented by a ``.'' followed by the characters.
For example, \texttt{.hello} is
a representation of the atom ``hello''.  A special character can be
represented by \texttt{.0xXX}, where \texttt{XX} is the hexidecimal
unicode for the character.
Atoms should not be confused with strings.
There are no special operations on atoms.
(In the future, operators may be introduced that convert between strings
and atoms.)

\section{Set}

In Harmony you can create a set of any collection of Harmony values.
Its syntax is \texttt{\{ $v_0, v_1, ...$ \}}.
Python users: note that in Harmony the empty set is denoted as $\{\}$.

The \texttt{set} module (\autoref{ap:set})
contains various convenient routines that operate on sets.
Native operations on sets include:

\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{len $e$} & cardinality \\
\hline
\texttt{$e$ - $e$} & set difference \\
\hline
\texttt{$e$ \& $e$ \& ...} & intersection \\
\hline
\texttt{$e$ | $e$ | ...} & union \\
\hline
\texttt{$e$ \string^ $e$ \string^ ...} & inclusion/exclusion (elements in odd number of sets) \\
\hline
\texttt{choose $e$} & select an element (Harmony will try all) \\
\hline
\texttt{min $e$} & minimum element \\
\hline
\texttt{max $e$} & maximum element \\
\hline
\end{tabular}
\end{center}

\section{Dictionary}

A dictionary maps a set of values (known as \emph{keys})
to another set of values.
The generic syntax of a dictionary is
\texttt{dict\{ $k_0:v_0, k_1:v_1, ...$ \}}.
If there are duplicate keys in the list, then only the one with
the maximum value survives.
Therefore the order of the keys in the dictionary does not matter.

There are various special cases of dictionaries,
including lists, tuples, strings,
and bags (multisets) that are individually described below.

Operations on dictionaries include the following:

\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{$d~k$} & indexing \\
\hline
\texttt{len $d$} & the number of keys in $d$ \\
\hline
\texttt{keys $d$} & the set of keys in $d$ \\
\hline
\texttt{$v$ [not] in keys $d$} & check if $v$ is a \emph{key} in $d$ \\
\hline
\texttt{$v$ [not] in $d$} & check if $v$ is a \emph{value} in $d$ \\
\hline
\texttt{$d$ \& $d$ \& ...} & dictionary intersection \\
\hline
\texttt{$d$ | $d$ | ...} & dictionary union \\
\hline
\end{tabular}
\end{center}

Because in Harmony brackets are used for parsing purposes only,
you can write $d[k]$ (or $d(k)$) instead of $d~k$.
However, if $k$ is an atom, like \texttt{.id}, then you might
prefer the notation $k\mathtt{.id}$.

Python users beware: the Harmony \texttt{$v$ in $d$} operator checks
if there is some key $k$ such that $d[k] = v$.  In Python, the same
syntax checks if $v$ is a key in $d$.  The difference exists because
in Harmony a list is a special case of a dictionary.

Dictionary intersection and dictionary union are defined so that
they work well with bags.  With disjoint dictionaries, intersection
and union work as expected.  If there is a key in the intersection, then
dictionary intersection retains the minimum value while dictionary
union retains the maximum value.  Unlike Python, Harmony
dictionary intersection and union are commutative and associative.

\section{List or Tuple}

In Harmony, there is no distinction between a list or a tuple.
Both are a special case of dictionary.  In particular, a list of
$n$ values is represented by a dictionary that maps the integers
$0, ..., n-1$ to those values.  Hence, if $t$ is a list or tuple,
then the notation $t[0]$ returns the first element of the list.

You can denote a list by a sequence of values, each value terminated
by a comma.  As per usual, you can use brackets or parentheses
at your discretion.  For Python users, the important thing to note
is that a singleton lists in Harmony must contain a comma.  For
example $[1,]$ is a list containing the value $1$, while $[1]$
is simply the value $1$.

The \texttt{list} module (\autoref{ap:list}) contains various
convenient routines that operate on lists or tuples.
Native operations on lists or tuples include the following:

\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{$t~k$} & indexing \\
\hline
\texttt{$t$ + $t$ + ...} & concatenation \\
\hline
\texttt{$t$ * $n$} & $n$ copies of $t$ concatenated \\
\hline
\texttt{$v$ [not] in $t$} & check if $v$ is a \emph{value} in $t$ \\
\hline
\texttt{len $t$} & the length of $t$ \\
\hline
\texttt{$t$ \& $t$ \& ...} & pairwise list minimum \\
\hline
\texttt{$t$ | $t$ | ...} & pairwise list maximum \\
\hline
\end{tabular}
\end{center}

\section{String}

In Harmony, a string is a list of single-character atoms.
The Python string notations mostly work in Harmony as well.
For example, \texttt{'abc'} is a three-character string
consisting of the atoms \texttt{.a}, \texttt{.b}, and
\texttt{.c}.
\texttt{"abc'} is the same three-character string.
Multi-line strings can be terminated by triple quotes or
triple double-quotes.

\section{Bag or Multiset}

Harmony has native support for bags (aka multisets), which are sets
that can contain multiple copies of the same element.
It is represented by a dictionary that maps each element to its
multiplicity.
So \texttt{bag\{ 10, 10, 12 \} = dict\{ 10: 2, 12: 1 \}}.

The \texttt{bag} module (\autoref{ap:bag}) contains various
convenient routines that operate on lists or tuples.
Native operations on bags include the following:

\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{$v$ [not] in keys $b$} & check if $v$ is in $b$ \\
\hline
\texttt{$t$ \& $t$ \& ...} & bag intersection\\
\hline
\texttt{$t$ | $t$ | ...} & bag union \\
\hline
\end{tabular}
\end{center}

\section{Program Counter}

A program counter is an integer that can be used to index into Harmony
bytecode.  When you define a method or a lambda function, you are
creating a constant of the program counter type.
Note, however, that labels are currently just integers.  This may
change in the future.

\section{Address}

In Harmony, each shared variable has an address, which is essentially
a list of Harmony values.
For example, the address of variable \texttt{d.v[3]} consists of the
list \texttt{.d}, \texttt{.v}, and 3.  The only way to construct an
address in Harmony is using the \texttt{?} operator.
\texttt{?d.v[3]} returns the address of variable \texttt{d.v[3]}.
An address can be dereferenced to return the value of the variable.
If $a$ is an address, then $!a$ returns the value.
Like C, Harmony supports the shorthand \texttt{$a$->v} for 
the expression \texttt{($!a$).v}.

\section{Context}

A context value captures the state of a thread.
A context is itself composed over various Harmony values.
Some of these values can be extracted.  In particular, if $c$
is a context, then:

\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{\texttt{c}.name} & returns the name of the thread's main method \\
\hline
\texttt{\texttt{c}.entry} & returns the program counter of the thread's main method \\
\hline
\texttt{\texttt{c}.arg} & returns the argument value of the thread's main method \\
\hline
\texttt{\texttt{c}.this} & returns the thread-local state \\
\hline
\texttt{\texttt{c}.mode} & returns the \emph{mode} of the context \\
\hline
\end{tabular}
\end{center}

The modes of a context are as follows:

\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{normal} & context of a executing thread \\
\hline
\texttt{stopped} & context of a stopped thread \\
\hline
\texttt{terminated} & context of a terminated thread \\
\hline
\texttt{failed} & context of a failed thread \\
\hline
\end{tabular}
\end{center}

@@

{\small
\begin{tabular}{|l|l|}
\hline
\texttt{$e$ == $e$}, \texttt{$e$ != $e$} & two Harmony values of the same type \\
\hline
\texttt{$e$ < $e$, $e$ <= $e$, $e$ > $e$, $e$ >= $e$} & any two Harmony values \\
\hline
\texttt{$e$ and $e$ and ...} & two or more booleans \\
\hline
\texttt{$e$ or $e$ or ...} & two or more booleans \\
\hline
\texttt{$e$ => $e$} & two booleans (implication) \\
\hline
\texttt{not $e$} & a boolean \\
\hline
\texttt{-$e$, \string~$e$, abs($e$)} & an integer \\
\hline
\texttt{$e$ + $e$ + ...} & two or more integers, tuples, or lists \\
\hline
\texttt{$e$ - $e$} & two integers or sets \\
\hline
\texttt{$e$ \& $e$ \& ..., $e$ | $e$ | ..., $e$ \string^ $e$ \string^ ...} & two or more integers or sets \\
\hline
\texttt{$e$ * $e$} & two integers or an integer and a list \\
\hline
\texttt{$e$ / $e$, $e$ // $e$} & two integers (integer division) \\
\hline
\texttt{$e$ \% $e$, $e$ mod $e$} & two integers (division remainder) \\
\hline
\texttt{$e$ ** $e$, $e << e$, $e >> e$} & two integers \\
\hline
\texttt{$\{ e..e \}$} & two integers \\
\hline
\texttt{$e$ [not] in $e$} & first is any Harmony value, second is a set or dict \\
\hline
\texttt{$e$ if $e$ else $e$} & middle $e$ must be a boolean \\
\hline
\texttt{!$e$} & an address \\
\hline
\texttt{choose $e$} & a set \\
\hline
\texttt{min $e$, max $e$, any $e$, all $e$, len $e$} & a set or a dict \\
\hline
\texttt{keys $e$} & a dictionary (includes tuple and list) \\
\hline
\texttt{atLabel $e$} & atom (corresponding to a label) \\
\hline
\texttt{nametag $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{setintlevel} \textit{e} & \textit{e} is a boolean \\
\hline
\texttt{stop} \textit{lv} & \textit{lv} is an lvalue \\
\hline
\texttt{?}\textit{lv} & \textit{lv} is an lvalue \\
\hline
\texttt{lambda $a$: $e$ end} & $a$ is a bounded variable, $e$ an expression\\
\hline
\end{tabular}
}
\vspace{1em}

@@

\section{The \texttt{bag} module}
\label{ap:bag}

@@

\section{The \texttt{set} module}
\label{ap:set}
